Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-02-25 application.tex
minor changes, new title

\subsection{Applicability of the VSI to the MMSN}
To obtain an overall picture of the relevance of the VSI to the MMSN,
we calculate characteristic growth times $t_\mathrm{grow} = 1/\nu$ of 
the fundamental mode in units of the local Keplerian orbital period
$P_\mathrm{orb}= 2\pi/\Omega_k$. The results are shown in 
Fig. \ref{mmsn_overall} for perturbation wavenumbers $5\lesssim
\khat\lesssim 50$. We consider this range because for  
$\khat=O(1)$ or $\khat=O(10^2)$ we found eigenfunctions have large
amplitudes near the vertical boundaries, so such modes may be affected
by boundary conditions.

In the MMSN the fundamental VSI operates at tens of AU, with
characteristic growth times of tens of orbits. %In our example, the VSI
% is most favored at $r\simeq 20\mathrm{AU}$ with a growth time
% $t_\mathrm{grow}\simeq 30$ orbits.
Fig. \ref{mmsn_overall} indicates that with decreasing radius, the VSI
shifts to smaller scales (larger $\khat$), since the disk becomes optically
thick so only small-scale disturbances have sufficiently short thermal
relaxation timescales. At larger radii ($r\gtrsim 
40\mathrm{AU}$,) the VSI can operate on larger scales (smaller
$\khat$) and growth times do not depend strongly on
$\khat$, since the disk is optically thin. However, approaching
$r\simeq 100\mathrm{AU}$ growth times become long, at $O(100)$ orbits so
the VSI may not be dynamically important in the outermost parts of the
MMSN.

>>>>

\subsection{VSI in the MMSN}
To obtain an overall picture of the relevance of the VSI to the MMSN,
we calculate characteristic growth times $t_\mathrm{grow} = 1/\nu$ of 
the fundamental mode in units of the local Keplerian orbital period
$P_\mathrm{orb}= 2\pi/\Omega_k$. The results are shown in 
Fig. \ref{mmsn_overall} for perturbation wavenumbers $5\lesssim
\khat\lesssim 50$. We consider this range because for  
$\khat=O(1)$ or $\khat=O(10^2)$ we found eigenfunctions have large
amplitudes near the vertical boundaries, so such modes may be affected
by boundary conditions.

In the MMSN the fundamental VSI operates at tens of AU, with
characteristic growth times of tens of orbits. %In our example, the VSI
% is most favored at $r\simeq 20\mathrm{AU}$ with a growth time
% $t_\mathrm{grow}\simeq 30$ orbits.
Fig. \ref{mmsn_overall} indicates that with decreasing radius, the VSI
shifts to smaller scales (larger $\khat$), since the disk becomes optically
thick so only small-scale disturbances have sufficiently short thermal
relaxation timescales. At larger radii ($r\gtrsim 
40\mathrm{AU}$), the disk is optically thin so growth times do not depend strongly on   
$\khat$, and the VSI can operate on larger scales (smaller $\khat$).  
However, approaching 
$r\simeq 100\mathrm{AU}$ growth times become long at $O(10^2)$ local orbits,
so the VSI may not be dynamically important in the outermost parts of
the MMSN.

### 2015-02-25 linear_adia.tex
minor changes, new title

Here we estimate the maximum thermal relaxation timescale
$\beta$ that allows the fundamental VSI for $\khat\gg1$.   
% Notice in Fig. \ref{relax_bcrit} the upper limit on $\beta$ for the
% fundamental isothermal VSI tends to a constant for large
% $\khat$. Using this clue, we can calculate this critical $\beta$ by
% considering $\khat\to\infty$. We proceed as follows. 
At marginal stability, the eigenfrequency $\hat{\sigma}=\hat{\omega}$
is real. The dispersion relation for the fundamental mode is then
\begin{align}\label{relax_disp_fund}
  0 = \sum_{l=1}^{6}c_l\hat{\omega}^l \quad \text{for $M=\hat{\nu}=0$}.  
\end{align} 
Taking the real and imaginary parts of Eq. \ref{relax_disp_fund} and considering 
$\khat\gg1$, we find
\begin{align}
  0 =& \left[1 + \gamma\beta^2\left(1-\gamma\right)\right] + 2\epsilon q
  \gamma\beta (\hat{\omega}\khat) -  (\hat{\omega}\khat)^2 \notag\\
  &+ \beta^2\gamma^2\hat{\omega}^2 (\hat{\omega}\khat)^2,\label{relax_cond1}\\
  0=& \beta(1-\gamma)\khat^2 + \epsilon q \khat^2 (\hat{\omega}\khat)
  - 2\beta (\hat{\omega}\khat)^2 - \epsilon q \gamma^2\beta^2 (\hat{\omega}\khat)^3.\label{relax_cond2}
\end{align}
This is a pair of simultaneous equations for
$\hat{\omega}$ and $\beta$.%  A simple solution can be obtained by the
% following considerations.
Now, our discussion of stable solutions 
without vertical shear yield a dispersion
relation $\hat{\omega}\propto \khat^{-1}$ for $\khat\gg 1$
(\S\ref{iso_stable}, Eq. \ref{iso_stable_disp}). We  therefore 
seek solutions to Eq. \ref{relax_cond1}---\ref{relax_cond2} with 
%we wish to seek a criterion independent of k 
%\begin{align}
%  \khat \to \infty, \quad \hat{\omega}\to 0 ,\quad \hat{\omega} \khat\text{ finite}.
%\end{align}
\begin{align}
  \khat \gg1, \quad |\hat{\omega}|\ll 1 ,\quad
  |\hat{\omega}\khat|=O(1). 
\end{align}
Then
\begin{align}
  \hat{\omega}\khat = \frac{(\gamma-1)\beta}{\epsilon q}
\end{align}
from Eq. \ref{relax_cond2}, which implies, from Eq. \ref{relax_cond1}
that
\begin{align}
  \beta = \frac{1}{(\gamma-1)}\left[\frac{1}{\left(\epsilon
        q\right)^2} - \frac{\gamma}{(\gamma-1)}\right]^{-1/2} 
\end{align}
at marginal stability for perturbations with large $\khat$. 
For a thin disk $\epsilon\ll 1$, we infer 
\begin{align}\label{iso_vsi_cond}
  \beta \lesssim \frac{|\epsilon q|}{\gamma-1} \equiv
  \beta_\mathrm{crit} 
\end{align}
is required for the fundamental VSI to operate. From our discussion
for isothermal perturbations (\S\ref{iso_discuss}), we expect
$\beta<\beta_\mathrm{crit}$ to be a sufficient condition for the
VSI. That is, when $\beta<\beta_\mathrm{crit}$, perturbations are
effectively isothermal. For a thin disk 
$\beta_\mathrm{crit}\ll 1$, which suggests that finite thermal
relaxation rapidly stabilizes the VSI.

>>>>

Here we estimate the maximum thermal relaxation timescale
$\beta$ that allows the fundamental VSI for $\khat\gg1$.   
% Notice in Fig. \ref{relax_bcrit} the upper limit on $\beta$ for the
% fundamental isothermal VSI tends to a constant for large
% $\khat$. Using this clue, we can calculate this critical $\beta$ by
% considering $\khat\to\infty$. We proceed as follows. 
At marginal stability, the eigenfrequency $\hat{\sigma}=\hat{\omega}$
is real. The dispersion relation for the fundamental mode is then
\begin{align}\label{relax_disp_fund}
  0 = \sum_{l=1}^{6}c_l\hat{\omega}^l \quad \text{for $M=\hat{\nu}=0$}.  
\end{align} 
Taking the real and imaginary parts of Eq. \ref{relax_disp_fund} and considering 
$\khat\gg1$, we find
\begin{align}
  0 =& \left[1 + \gamma\beta^2\left(1-\gamma\right)\right] + 2\epsilon q
  \gamma\beta (\hat{\omega}\khat) -  (\hat{\omega}\khat)^2 \notag\\
  &+ \beta^2\gamma^2\hat{\omega}^2 (\hat{\omega}\khat)^2,\label{relax_cond1}\\
  0=& \beta(1-\gamma)\khat^2 + \epsilon q \khat^2 (\hat{\omega}\khat)
  - 2\beta (\hat{\omega}\khat)^2 - \epsilon q \gamma^2\beta^2 (\hat{\omega}\khat)^3.\label{relax_cond2}
\end{align}
This is a pair of simultaneous equations for
$\hat{\omega}$ and $\beta$. % A simple solution can be obtained by the
% following considerations.
 Our discussion of stable solutions 
without vertical shear yield a dispersion
relation $\hat{\omega}\propto \khat^{-1}$ for $\khat\gg 1$
(\S\ref{iso_stable}, Eq. \ref{iso_stable_disp}). We  therefore 
seek solutions to Eq. \ref{relax_cond1}---\ref{relax_cond2} with 
%we wish to seek a criterion independent of k 
%\begin{align}
%  \khat \to \infty, \quad \hat{\omega}\to 0 ,\quad \hat{\omega} \khat\text{ finite}.
%\end{align}
\begin{align}
  \khat \gg1, \quad |\hat{\omega}|\ll 1 ,\quad
  |\hat{\omega}\khat|=O(1). 
\end{align}
Then
\begin{align}
  \hat{\omega}\khat = \frac{(\gamma-1)\beta}{\epsilon q}
\end{align}
from Eq. \ref{relax_cond2}, which implies, from Eq. \ref{relax_cond1}
that
\begin{align}
  \beta = \frac{1}{(\gamma-1)}\left[\frac{1}{\left(\epsilon
        q\right)^2} - \frac{\gamma}{(\gamma-1)}\right]^{-1/2} 
\end{align}
at marginal stability for perturbations with large $\khat$. 
For a thin disk $\epsilon\ll 1$, we infer 
\begin{align}\label{iso_vsi_cond}
  \beta \lesssim \frac{|\epsilon q|}{\gamma-1} \equiv
  \beta_\mathrm{crit} 
\end{align}
is required for the fundamental VSI to operate. From our discussion
for isothermal perturbations (\S\ref{iso_discuss}), we expect
$\beta<\beta_\mathrm{crit}$ to be a sufficient condition for the
VSI. That is, when $\beta<\beta_\mathrm{crit}$, perturbations are
effectively isothermal. For a thin disk 
$\beta_\mathrm{crit}\ll 1$, which suggests that finite thermal
relaxation rapidly stabilizes the VSI.
